---
title: "KY_2021_Tornado"
output: html_document
date: "2026-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

##Packages
```{r message=FALSE}

library(tidyverse)
library(foreign)
library(broom)
library(readxl)
library(plm)
library(tidycensus)
library(MatchIt)
library(countyhealthR)
library(data.table)

```

##ACS Data 

Matching
```{r}

#Load ACS Data Using Tidy Census
Income <- get_acs(
geography = "county",
variables = "B19013_001",# Household Income
state = "KY",
output = "wide", # creates output with one row for each county
#geometry = TRUE, When wanting column for GIS work
year = 2021 #Using 2021 as our baseline year for comparison and matching
)


# Split the 'items' column into multiple new columns (e.g., up to 3)
# Note: separate_wider_delim is recommended for newer tidyr versions (1.3.0+)
Income <- Income %>%
  separate_wider_delim(
    NAME,
    delim = ",",
    names = c("County", "State"),
    too_few = "align_start" # Aligns items to the start, filling missing with NA
  )

#Remove County from Row for matching 
Income <- Income %>%
  separate_wider_delim(
    County,
    delim = " ",
    names = c("County", "Del"),
    too_few = "align_start" # Aligns items to the start, filling missing with NA
  )

#ACS Data with Population by County
Cty_pop <- get_acs(
geography = "county",
variables = "B01003_001",# Population By County
state = "KY",
output = "wide", # creates output with one row for each county
#geometry = TRUE, When wanting column for GIS work
year = 2021
)

# Split the 'items' column into multiple new columns (e.g., up to 3)
# Note: separate_wider_delim is recommended for newer tidyr versions (1.3.0+)
Cty_pop <- Cty_pop %>%
  separate_wider_delim(
    NAME,
    delim = ",",
    names = c("County", "State"),
    too_few = "align_start" # Aligns items to the start, filling missing with NA
  )

#Remove County from Row for matching
Cty_pop <- Cty_pop %>%
  separate_wider_delim(
    County,
    delim = " ",
    names = c("County", "Del"),
    too_few = "align_start" # Aligns items to the start, filling missing with NA
  )

#Merge DFs and remove the County Extra Column
cty_match <- left_join(Cty_pop, Income, by = c("GEOID", "County", "State", "Del")) 
cty_match <- select(cty_match, -Del)

#Add Tornado "Treatment Variable" for matching
cty_match <- cty_match %>%
  mutate(tornado = if_else(County == "Graves", 1,
                    if_else(County == "Fulton", 1,
                      if_else(County == "Hickman", 1,
                       if_else(County == "Marshall", 1,
                        if_else(County == "Lyon", 1, 
                          if_else(County == "Caldwell", 1,
                            if_else(County == "Hopkins", 1,
                              if_else(County == "Muhlenberg", 1, 
                                if_else(County == "Ohio", 1, 
                                  if_else(County == "Breckinridge", 1, 
                                    if_else(County == "Grayson", 1, 0)      
                              )      
                              )      
                              )      
                              )      
                              )      
                              )       
                              )        
                              )      
                              )       
                             ))



#Matching on Median Income and Mean Population
m.out <- matchit(tornado ~ B01003_001E + B19013_001E, data = cty_match, ratio = 2 ,method = "nearest")
summary(m.out, un = FALSE) # un = FALSE shows only matched results
matched_data <- match.data(m.out)


# Convert the column to a comma-separated string
match_string <- toString(matched_data$County)

# Print the result
print(match_string)
# Output: [1] "apple, banana, orange, grape"


rm(cty_match)
rm(m.out)
rm(Cty_pop)
rm(Income)
rm(match_string)

#Save Data

#acs_match <- readRDS(matched_data, file = "C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/ACS_KY/match_ky.rds")

```


##KIDS Count Data Load & Management
```{r message=FALSE}

#child_poverty_rate_single_year_estimates: Percent of children living below the official federal poverty level.
c_pov <- read_excel("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/Kids_Count/Child poverty rate (single-year estimates).xlsx")

c_pov <- c_pov %>%
  rename(County = Location, Percent_Poverty = Data, Year = TimeFrame)

#Children_living_in_food_insecure_households: Percentage of children who live in households that at times lack access to enough food for healthy living or unreliable availability to nutritionally adequate foods. 
c_food <- read_excel("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/Kids_Count/Children living in food insecure households (2018 and beyond).xlsx")

c_food <- c_food %>%
  rename(County = Location, Percent_FoodInsecure = Data, Year = TimeFrame)

#Children_receiving_aid_from_Supplemental_Nutrition_Assistance_Program_monthly_average: Average monthly number of children under 18 who participated in the Supplemental Nutrition Assistance Program (SNAP), formerly called food stamps. 
c_snap <- read_excel("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/Kids_Count/Children receiving aid from Supplemental Nutrition Assistance Program (monthly average).xlsx")

c_snap <- c_snap %>%
  rename(County = Location, Avg_number_snap = Data, Year = TimeFrame)

#Infants_and_children_receiving_Women_Infants_and_Children: Average monthly number of infants and children up to age 5 served by the Special Supplemental Nutrition Program for Women, Infants, and Children (WIC).
 
c_wic <- read_excel("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/Kids_Count/Infants and children receiving Women, Infants, and Children assistance (monthly average).xlsx")

c_wic <- c_wic %>%
  rename(County = Location, Avg_number_wic = Data, Year = TimeFrame)

#Join All KIDS Count
kids_count <- c_food %>%
  left_join(c_pov, by =c("LocationType", "County", "Year")) %>%
  left_join(c_snap, by =c("LocationType", "County", "Year")) %>%
  left_join(c_wic, by =c("LocationType", "County", "Year"))

kids_count <- select(kids_count, -DataFormat.x)
kids_count <- select(kids_count, -DataFormat.y)
kids_count <- select(kids_count, -DataFormat.x.x)
kids_count <- select(kids_count, -DataFormat.y.y)


rm(c_food)
rm(c_pov)
rm(c_snap)
rm(c_wic)

# Save the data frame as an RDS file
#saveRDS(kids_count, file = "C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/kids_count.rds")
```


```{r}

kids_count <- readRDS("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/Kids_Count/kids_count.rds")

```

##RWJF

"Ballard", "Bath", "Bourbon", "Boyle", "Bracken", "Breckinridge", "Caldwell", "Carlisle", "Carter", "Edmonson", "Floyd", "Fulton", "Gallatin", "Grant", "Graves", "Grayson", "Hancock", "Hickman", "Hopkins", "Knox", "Lincoln", "Logan", "Lyon", "Marshall", "Meade", "Muhlenberg", "Nelson", "Ohio", "Owsley", "Perry", "Shelby", "Wayne", "Whitley"

Data Load & Reduction
```{r}

#Load Data
chr_2018 <- read_csv("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/analytic_data2018_0.csv")
colnames(chr_2018) <- tolower(colnames(chr_2018))

#dput(names(chr_2018)) List headers with ""

chr_2019 <- read_csv("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/analytic_data2019.csv")
colnames(chr_2019) <- tolower(colnames(chr_2019))

chr_2020 <- read_csv("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/analytic_data2020_0.csv")
colnames(chr_2020) <- tolower(colnames(chr_2020))

chr_2021 <- read_csv("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/analytic_data2021.csv")
colnames(chr_2021) <- tolower(colnames(chr_2021))

chr_2022 <- read_csv("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/analytic_data2022.csv")
colnames(chr_2022) <- tolower(colnames(chr_2022))

chr_2023 <- read_csv("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/analytic_data2023_0.csv")
colnames(chr_2023) <- tolower(colnames(chr_2023))

chr_2024 <- read_csv("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/analytic_data2024.csv")
colnames(chr_2024) <- tolower(colnames(chr_2024))

#Match on FIPS/GEOID

#Find Match Columns with intersect
#RBind on only columns that are in the intersected list

#2018|2019
InBoth <- intersect(colnames(chr_2018), colnames(chr_2019))
chr_comb <- rbind(chr_2018[,InBoth], chr_2019[,InBoth])

#2020

InBoth <- intersect(colnames(chr_comb), colnames(chr_2020))
chr_comb <- rbind(chr_comb[,InBoth], chr_2020[,InBoth])

#2021

InBoth <- intersect(colnames(chr_comb), colnames(chr_2021))
chr_comb <- rbind(chr_comb[,InBoth], chr_2021[,InBoth])

#2022

InBoth <- intersect(colnames(chr_comb), colnames(chr_2022))
chr_comb <- rbind(chr_comb[,InBoth], chr_2022[,InBoth])

#2023

InBoth <- intersect(colnames(chr_comb), colnames(chr_2023))
chr_comb <- rbind(chr_comb[,InBoth], chr_2023[,InBoth])

#2024

InBoth <- intersect(colnames(chr_comb), colnames(chr_2024))
chr_comb_final <- rbind(chr_comb[,InBoth], chr_2024[,InBoth])

#saveRDS(chr_comb_final, file = "C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/chr_ky_2018_2024.rds")

```


```{r}

chr_ky_1824 <- readRDS("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/chr_ky_2018_2024.rds")

```


#Data Merge
```{r}

##ACS Data
acs_match <- readRDS("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/ACS_KY/match_ky.rds")
colnames(acs_match) <- tolower(colnames(acs_match))
#Extract Matched Columns
acs <- acs_match %>% select("geoid", "county")

rm(acs_match)

#RWJF Data
chr_ky_1824 <- readRDS("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/CHR_2018_2024/chr_ky_2018_2024.rds")
colnames(chr_ky_1824) <- tolower(colnames(chr_ky_1824))


#Merge DFs and remove the County Extra Column
comb_df <- left_join(chr_ky_1824, acs, by = c("geoid")) 

comb_df$county <- as.factor(comb_df$county)
comb_df$year <- as.factor(comb_df$year)


##KIDS Count Data
kids_count <- readRDS("C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/Kids_Count/kids_count.rds")
colnames(kids_count) <- tolower(colnames(kids_count))

kids_count$county <- as.factor(kids_count$county)
kids_count$year <- as.factor(kids_count$year)

#Merge DFs and remove the County Extra Column
comb_final <- left_join(kids_count, comb_df, by = c("year", "county"))


values_to_keep <- c("Ballard", "Bath", "Bourbon", "Boyle", "Bracken", "Breckinridge", "Caldwell", "Carlisle", "Carter", "Edmonson", "Floyd", "Fulton", "Gallatin", "Grant", "Graves", "Grayson", "Hancock", "Hickman", "Hopkins", "Knox", "Lincoln", "Logan", "Lyon", "Marshall", "Meade", "Muhlenberg", "Nelson", "Ohio", "Owsley", "Perry", "Shelby", "Wayne", "Whitley")
filtered_data <- comb_final %>%
  filter(county %in% values_to_keep)

#dput(names(filtered_data)) #List headers with ""

# Remove columns that contain the word "Numerator and Denominator"
filtered_data <- filtered_data %>%
  select(-contains("numerator"))

filtered_data <- filtered_data %>%
  select(-contains("denominator"))

# Remove columns B and D
filtered_data <- select(filtered_data, -geoid, -name, -locationtype, -"state abbreviation")

#saveRDS(filtered_data, file = "C:/Users/Custom/Dropbox/Post Doctoral Work/Pittsburgh_MCH/Kentucky_Tornado_2021_Disaster/Data/ky_tornado_21_data.rds")

```